---
title: "Correcting Associations for Measurement Overlap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correcting Associations for Measurement Overlap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

```

```{r setup}
library(latentoverlap)
```

# Why correct for measurement overlap?

When two constructs are measured with partly overlapping items, their observed total-score correlation may partly reflect shared item content rather than a construct-level association. The latentoverlap package re-expresses the observed total-score correlation as the latent association implied by an explicit measurement model.

The corrected association should therefore be interpreted as model-based: it depends on the assumed reliabilities, loadings, thresholds, and degree of overlap.

# Basic continuous example

Suppose two six-item scales correlate r = .60. Both scales have reliabilities around .80, and one item in each scale is assumed to overlap with a cross-loading of .30.

```{r}

mod <- reliability_model(
  relA = .80,
  relB = .85,
  kA = 6,
  kB = 6
)

rho_correction(
  r_obs = .60,
  type = "continuous",
  measurement_model = mod,
  cA = .30,
  cB = .30,
  k_overlap = 1,
  ci = "none"
)
```


## Analytic confidence intervals

The recommended default is the analytic confidence interval.

### Using standard error observed association
```{r}
rho_correction(
  r_obs = .60,
  type = "continuous",
  measurement_model = mod,
  cA = .30,
  cB = .30,
  k_overlap = 1,
  ci = "analytic",
  r_SE = .10
)
```

### Using confidence intervals observed association
```{r}
rho_correction(
  r_obs = .60,
  type = "continuous",
  measurement_model = mod,
  cA = .30,
  cB = .30,
  k_overlap = 1,
  ci = "analytic",
  r_ci = c(.45, .72)
)

```

# Ordinal item scores example

For ordinal data, thresholds must be specified. Ordinal corrections are computationally more intensive. The analytic CI is therefore typically preferred over bootstrap.

```{r}
thr5 <- c(-1.645, -0.643, 0.643, 1.645)

rho_correction(
  r_obs = .60,
  type = "ordinal",
  measurement_model = mod,
  cA = .30,
  cB = .30,
  k_overlap = 1,
  thrA = thr5,
  thrB = thr5,
  ci = "analytic",
  r_SE = .10
)
```


# Parametric bootstrapped confidence interval

The bootstrap uses the same Fisher-z uncertainty model but is slower, especially for ordinal models.

```{r}
rho_correction(
  r_obs = .60,
  type = "continuous",
  measurement_model = mod,
  cA = .30,
  cB = .30,
  k_overlap = 1,
  ci = "bootstrap",
  r_SE = .10,
  n_boot = 500
)
```


# Sensitivity analysis
The corrected association depends on the assumed overlap. A sensitivity analysis is therefore recommended.

## Multiply overall overlap degree across grid values

```{r}
sens <- rho_overlap_sensitivity(
  r_obs = .60,
  type = "continuous",
  measurement_model = mod,
  cA = .30,
  cB = .30,
  k_overlap = 1,
  ci = "analytic",
  r_SE = .10,
  grid = seq(0, 2, by = .05)
)

plot(sens)
```

overlap_value = 1 corresponds to the assumed overlap
Values below/above represent weaker/stronger overlap


## Vary overlap degree in item-specific sensitivity analysis

Instead of varying all overlap parameters simultaneously, a sensitivity analysis can also evaluate how strongly the corrected association depends on one specific overlapping item.

In the example below, only the first overlapping item in scale A is varied while the remaining overlap assumptions are kept constant.

```{r}
sens_item <- rho_overlap_sensitivity(
  r_obs = .60,
  type = "continuous",
  measurement_model = mod,
  cA_vec = c(.30, 0, 0, 0, 0, 0),
  cB_vec = c(.30, 0, 0, 0, 0, 0),
  ci = "analytic",
  r_SE = .10,
  vary = "item",
  focal_scale = "A",
  focal_item = 1,
  grid = seq(0, .80, by = .05),
  show_progress = FALSE
)

head(sens_item$results)
```

The sensitivity curve can again be visualized using the plot method.

```{r}
plot(sens_item)
```

This type of analysis is useful when researchers suspect that one specific overlapping item may disproportionately influence the corrected association.



# Reporting results

A transparent report should include:

- observed total-score correlation
- assumed measurement model
- overlap assumptions
- corrected association and CI
- sensitivity analysis

## Example:
The observed correlation was corrected using a parallel measurement model with one overlapping item pair (cross-loadings = .30). The corrected latent association was reported with an analytic confidence interval. A sensitivity analysis evaluated robustness across alternative overlap assumptions.

